# Learning Check Solutions {#appendixD}

```{r setup_lc_solutions, include=FALSE, purl=FALSE}
opts_chunk$set(tidy = FALSE, out.width = "\\textwidth")
# This bit of code is a bug fix on asis blocks, which we use to show/not show LC solutions, which are written like markdown text. In theory, it shouldn't be necessary for knitr versions <=1.11.6, but I've found I still need to for everything to knit properly in asis blocks. More info here:
# https://stackoverflow.com/questions/32944715/conditionally-display-block-of-markdown-text-using-knitr
knit_engines$set(asis = function(options) {
  if (options$echo && options$eval) knit_child(text = options$code)
})
```

Here are the solutions to all Learning Checks throughout the book.



## Chapter 1 Solutions 

```{r, include=FALSE, purl=FALSE}
chap <- 1
lc <- 0
```

```{r message=FALSE}
library(dplyr)
library(ggplot2)
library(nycflights13)
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`**  Repeat the above installing steps, but for the `dplyr`, `nycflights13`, and `knitr` packages. This will install the earlier mentioned `dplyr` package, the `nycflights13` package containing data on all domestic flights leaving a NYC airport in 2013, and the `knitr` package for writing reports in R. 

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** "Load" the `dplyr`, `nycflights13`, and `knitr` packages as well by repeating the above steps.

**Solution**: If the following code runs with no errors, you've succeeded!

```{r, eval=FALSE}
library(dplyr)
library(nycflights13)
library(knitr)
```

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** What does any *ONE* row in this `flights` dataset refer to?

- A. Data on an airline 
- B. Data on a flight
- C. Data on an airport
- D. Data on multiple flights

**Solution**: This is data on a flight. Not a flight path! Example: 

* a flight path would be United 1545 to Houston
* a flight would be United 1545 to Houston at a specific date/time. For example: 2013/1/1 at 5:15am.


**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** What are some examples in this dataset of **categorical** variables?  What makes them different than **quantitative** variables?

**Solution**: Hint: Type `?flights` in the console to see what all the variables mean!

* Categorical:
    + `carrier` the company
    + `dest` the destination
    + `flight` the flight number. Even though this is a number, its simply a label. Example United 1545 is not  less than United 1714
* Quantitative:
    + `distance` the distance in miles
    + `time_hour` time

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** What properties of the observational unit do each of `lat`, `lon`, `alt`, `tz`, `dst`, and `tzone` describe for the `airports` data frame?  Note that you may want to use `?airports` to get more information.

**Solution**: `lat` `long` represent the airport geographic coordinates, `alt` is the altitude above sea level of the airport (Run `airports %>% filter(faa == "DEN")` to see the altitude of Denver International Airport), `tz` is the time zone difference with respect to GMT in London UK, `dst` is the daylight savings time zone, and `tzone` is the time zone label.

**`r paste0("(LC", chap, ".", (lc <- lc + 1), ")")`** Provide the names of variables in a data frame with at least three variables in which one of them is an identification variable and the other two are not.  In other words, create your own tidy dataset that matches these conditions.

**Solution**: 

* In the `weather` example in LC2.8, the combination of `origin`, `year`, `month`, `day`, `hour`  are identification variables as they identify the observation in question.
* Anything else pertains to observations: `temp`, `humid`, `wind_speed`, etc.



***

<!-- These need to be removed or replaced for the Biology version -->
